---
title: "Untitled"
output: html_document
date: "2023-07-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(stringr)

```

OBJECTIVE: Creating our own classes of incidents based on words present in comments column: 
SCOPE: in the year 2022
```{r}
df22 = df3 %>% filter(year == 2022)
```

DICTIONARIES 
```{r}
drug_words = c('[Dd]rug','[Ss]hooting up','[Pp]araphenelia','[Bg]aggie','[Nn]eedle','[Hh]igh', '[Hh]erion','[Mm]eth', '[Pp]ipe', '[Ii]ntoxicat', '[Cc]ocaine', '[Hh]ypodermic','[Ff]ume', '[Dd]ose','[Ff]entanyl','[Nn]arcotic','[Ii]nhalent', '[Cc]annabis','[Mm]arijuana','[Ff]oil',)


alcohol_words = c('[Bb]eer','[Aa]lcohol','[Dd]runk', "[Ii]ntoxicated")

df22 %>% filter(str_detect(comments, pattern = paste(drug_words, collapse = "|"))) 

df22 %>% filter(str_detect(comments, pattern = paste(alcohol_words, collapse = "|")))  





```

## words to put in dictionaries:
cussing and swearing, being verbally abusive
DO NOT BOARD 
refus
police 
physically 
assault
Disruption of service
aggressive
machettee 
axe
hachet
brandishing
gun
vaping
smoking
welfare chk
tent
attacking
transient
angry
crazy
verbal
disruptive
police


```{r}
df22 %>% filter(str_detect(comments, regex("do not board", ignore_case = TRUE)))



```




```{r}
sex_words = c('[Ll]ewd','[Ss]ex','[Mm]asturbat','[Ii]nappropriate','[Gg]enital','[Rr]ape','[Hh]arass')

df22 %>% filter(str_detect(comments, pattern = paste(sex_words, collapse = "|"))) %>%  group_by(tractce) %>% count() %>% arrange(desc(n))
```
```{r}
violent_words <- c('[Ss]wing', '[Hh]it', '[Ss]truck', '[Aa]ssault', '[Tt]hrow', '[Gg]rab',
           '[Tt]hreat', '[Pp]hysic', '[Mm]enac','[Ff]ace','[Ff]led', '[Ii]njur', '[Ss]teel pole', '[Ss]pit','[Ss]hov', '[Ss]lam','[Hh]urt', '[Bb]elig','[Pp]unch', '[Gg]oing off', '[Yy]ell','[Aa]ggressive','[Gg]un', '[Kk]nife','[Mm]achetee','[Aa]xe','[Br]andish','[Hh]achet','[Aa]xe','[Aa]ttack', '[Aa]ggres', '[Kk]ill', '[Pp]ound', '[Ss]uicid', '[Ff]ire', '[Ff]irearm', '[Hh]arras', '[Bb]lood')

df22 %>% filter(str_detect(comments, pattern = paste(violent_words, collapse = "|")))
```


```{r}
bio_words = c('[Uu]rin','[Bb]lood','[Bb]leed', '[Vv]omit','[Ss]ubstance', '[Bb]io harzard')

df22 %>% filter(str_detect(comments, pattern = paste(bio_words, collapse = "|")))
```

```{r}
library(tidyverse)
library(tidytext)
library(stringr)
library(MASS)

corpus <- Corpus(VectorSource(df22$comments))

corpus <- tm_map(corpus, content_transformer(tolower))
corpus <- tm_map(corpus, removePunctuation)
corpus <- tm_map(corpus, removeNumbers)
corpus <- tm_map(corpus, removeWords, stopwords("english"))
corpus <- tm_map(corpus, stemDocument)





```


```{r}
dtm <- DocumentTermMatrix(corpus)

library(topicmodels)

 LDA(dtm, k = 4, control = list(seed = 503))

topics = tidy(dtm_lda, matrix = "beta")

top_terms = topics %>% group_by(topic) %>% slice_max(beta, n = 10) %>% ungroup() %>% arrange(topic, -beta)

top_terms %>%
  mutate(term = reorder_within(term, beta, topic)) %>%
  ggplot(aes(beta, term, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ topic, scales = "free") +
  scale_y_reordered()


dtm_tidy %>% filter(!term %in% top_terms$term) %>% 
```
```{r}
corpus <- Corpus(VectorSource(df22$comments))

corpus <- tm_map(corpus, content_transformer(tolower))
corpus <- tm_map(corpus, removePunctuation)
corpus <- tm_map(corpus, removeNumbers)
corpus <- tm_map(corpus, removeWords, stopwords("english"))
corpus <- tm_map(corpus, stemDocument)

#corpus = tm_map(corpus, removeWords,top_terms$term )

dtm <- DocumentTermMatrix(corpus)
lda2  = LDA(dtm, k = 4, control = list(seed = 503))




```

```{r}
words  = df22 %>% unnest_tokens(word, comments) %>% 
  anti_join(stop_words) %>% 
  filter(!(word %in% (top_terms$term))) 

words <- words %>%
count(incident_id, word) %>%
group_by(incident_id) %>%
mutate(exists = (n>0)) %>%
ungroup %>%
group_by(word) %>%
mutate(total = sum(n)) %>%
filter(total > 100) %>%
pivot_wider(id_cols = incident_id, names_from = word, values_from = exists, values_fill = list(exists=0)) 

```
```{r}
library(broom)
kclust = kmeans(words, centers = 3)
kclust$centers

wink <- augment(kclust,words)
wink %>% group_by(.cluster) %>% count()

df22 = inner_join(df22, wink %>% dplyr::select(incident_id,.cluster), by = "incident_id")

df22 %>% filter(.cluster == 3)




```

```{r}
kclusts <- tibble(k = 1:9) %>%
mutate(
kclust = map(k, ~kmeans(words, .x)),
glanced = map(kclust, glance),
augmented = map(kclust, augment, words)
)

assignments <- kclusts %>%
unnest(augmented)

```

```{r}
clusterings <- kclusts %>%
unnest(glanced, .drop = TRUE)
ggplot(clusterings, aes(k, tot.withinss)) +
geom_line()
```

