---
title: "trimet_analysis_v2.2"
author: Karol Orozco, Corey Cassell, Justus Eaglesmith, Charles Hanks, & CorDarryl Hall
output: html_document
date: "2023-07-05"
---


Libraries: 
```{r}
library(tidyverse)
library(lubridate)
library(gridExtra)
library(forcats)
library(stringr)
library(tm)
library(SnowballC)
library(RColorBrewer)
library(tigris)
library(sf)
library(skimr)

library(gganimate)
library(lubridate)
library(ggmap)
```



Dataset: 
```{r}
df3 = read_rds('/Users/chanks/workspace/capstone/data/df_spatial_clean.rds')
```

## TIME ANALYSIS 
```{r}
# Group data by year only and count the number of incidents
df_grouped <- df3 %>%
  group_by(year) %>%
  summarise(total_incident_count = n(), .groups = "drop")

# Create ggplot object
p <- ggplot(df_grouped, aes(x = year, y = total_incident_count)) +
  geom_line(size = 1, color = "blue") +
  geom_point(data = subset(df_grouped, year == 2022), colour = "red", size = 4, shape = 4) +
  labs(
    title = "Change in Total Number of Incidents Over Time",
    x = "Year",
    y = "Total Number of Incidents"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_y_log10() +
  geom_text(
    data = subset(df_grouped, year == 2022),
    aes(label = ifelse(year == 2022, as.character(total_incident_count), "")),
    vjust = -2.5,
    hjust = 0.5,
    size = 4,
    colour = "red"
  )

# Add animation
p_animate <- p + transition_reveal(year)

# Render the animation
animate(p_animate, end_pause = 50, height = 400, width = 600)

```

Clearly significant results in the year 2022. This will be the year we focus on to prime our analysis. Lets look at by location per year.

# By Location
```{r}
# Group data by year and location, and count the number of incidents
df_grouped <- df3 %>%
  group_by(year, location) %>%
  summarise(incident_count = n(), .groups = "drop")

# Filter to top 10 locations by total incidents
top_locations <- df_grouped %>% 
  group_by(location) %>% 
  summarise(total_incidents = sum(incident_count), .groups = "drop") %>% 
  top_n(10, total_incidents) %>% 
  pull(location)

df_grouped <- df_grouped %>%
  filter(location %in% top_locations)

# Create ggplot object
p <- ggplot(df_grouped, aes(x = year, y = incident_count, color = location)) +
  geom_line(size = 1) +
  geom_point(data = subset(df_grouped, year == 2022), colour = "red", size = 4, shape = 4) +
  labs(
    title = "Change in Number of Incidents Over Time",
    x = "Year",
    y = "Number of Incidents",
    color = "Location"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_y_log10() +
  geom_text(
    data = subset(df_grouped, year == 2022),
    aes(label = ifelse(year == 2022, as.character(incident_count), "")),
    vjust = -.5,
    hjust = 0.5,
    size = 4,
    colour = "red"
  )
# Add animation
p_animate <- p + transition_reveal(year)

# Render the animation
animate(p_animate, end_pause = 50, height = 400, width = 600)

```
Appears that most of the incidents are occurring at Gateway, and Rose Quarter last year with some deviation between the top 3-5 following. Lets look to see if there is a trend with specific seasons/months.

```{r}
# Filter original dataframe to include only rows from the year 2022
df_2022 <- df3 %>%
  filter(year == 2022)

# Assign a 'season' column to df_2022
df_2022 <- df_2022 %>%
  mutate(season = case_when(
    month %in% c(12, 1, 2) ~ "Winter",
    month %in% 3:5 ~ "Spring",
    month %in% 6:8 ~ "Summer",
    month %in% 9:11 ~ "Autumn",
    TRUE ~ as.character(month)
  ))


# Assign a 'season' column to df_2022 based on 'month'
df_2022 <- df_2022 %>%
  mutate(season = case_when(
    month %in% c("Dec", "Jan", "Feb") ~ "Winter",
    month %in% c("Mar", "Apr", "May") ~ "Spring",
    month %in% c("Jun", "Jul", "Aug") ~ "Summer",
    month %in% c("Sep", "Oct", "Nov") ~ "Autumn",
    TRUE ~ as.character(month)
  ))
```

```{r}

# Calculate the number of incidents per month
df_sorted_2022 <- df_2022 %>%
   group_by(month, season) %>%
   summarise(incident_count = n(), .groups = "drop") %>%
   arrange(month)

# Calculate the average number of incidents per month
df_avg_2022 <- df_2022 %>%
  group_by(month) %>%
  summarise(avg_incident_count = n(), .groups = "drop") %>%
  arrange(month)

# Plotting the bar plot with the trendline
# Define a color palette
color_palette <- c("Winter" = "#1f78b4", "Spring" = "#33a02c", "Summer" = "#e31a1c", "Autumn" = "#ff7f00")

ggplot(df_sorted_2022, aes(x = month, y = incident_count, fill = season)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = color_palette) +
  labs(title = "Number of Incidents by Month in 2022", x = "Month", y = "Number of Incidents", fill = "Season") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "bottom") +
  geom_line(data = df_avg_2022, aes(x = month, y = avg_incident_count, group = 1), 
            color = "black", size = 1, inherit.aes = FALSE)
```

Appears to be a trend that in the warmer seasons incidents increase. Lets see if the statistics support this with a chi squared test of proportional independence.

```{r}

# Chi-square test for independence between month and incidents
chisq_test_month <- chisq.test(table(df_2022$month))

print(chisq_test_month)

# Chi-square test for independence between season and incidents
chisq_test_season <- chisq.test(table(df_2022$season))

print(chisq_test_season)

```
```{r}
# Extract standardized residuals
stdres <- chisq_test_month$stdres
names(stdres) <- month.abb

# Print residuals
print(stdres)

# Highlight residuals greater than 2 in absolute value
significant_months <- names(stdres)[abs(stdres) > 2]

print(paste("Months with significantly different number of incidents than expected: ", paste(significant_months, collapse = ", ")))

```

The following months show statistical significance, specifically in that April, September, and October have less than expected incidents while Jan, & May-July have more than expected incidents. Lets see if there are trends within time of day.


# GEO-SPATIAL ANALYSIS 

Justus' analysis of the months shows during the months of May through July, there were higher than expected security incidents. Let us examine where these events occured. 

```{r}
options(tigris_use_cache = TRUE)

this.year = 2020

mult_tracts = tracts(state = 'OR', county = "Multnomah", cb = T, year = this.year)
clack_tracts = tracts(state = 'OR', county = "Clackamas", cb = T, year = this.year)
wash_tracts = tracts(state = 'OR', county = "Washington", cb = T, year = this.year)

tri_county_tracts = bind_rows(mult_tracts,clack_tracts,wash_tracts)
```



Looking at where incidents occurred during peak period: May - July, 2022: 
```{r}
df_may_jul_2022 = df3 %>% filter(date >= "2022-01-05" & date <= "2022-07-31") %>% group_by(location_id) %>% summarize(n = n()) %>% arrange(desc(n))

df_may_jul_2022 = inner_join(df_may_jul_2022, df3 %>% select(location_id,stop_name, lon, lat, tractce, geometry.y) %>% distinct(), by = 'location_id')

#bubble plot: 

ggplot() +
  geom_sf(data = tri_county_tracts) +
  geom_point(data = df_may_jul_2022, aes(x = lon, y = lat, size = n, color = n, alpha = n)) + 
  theme(legend.position = "none") + 
  coord_sf(xlim = c(-123.1, -122.4), ylim = c(45.3, 45.7)) + 
  scale_color_gradient(low = 'red', high = 'darkred') + 
  labs(title = "Security incidents, May-Jul 2022")
```
Plotting with ggmap
```{r}
library(ggmap)
library(osmdata)

api_key = "AIzaSyBd4Q6cJzX1fFUhW2qgv6eWOztnZOcTuRk"
register_google(key = api_key)

portland_map = get_map(location = 'portland, or, usa', maptype = "roadmap", source = 'google', zoom = 10)

ggmap(portland_map) + 
  geom_point(data = df_may_jul_2022, aes(x = lon, y = lat, size = n, color = n, alpha = n)) + 
  theme(legend.position = "none") + 
  #coord_sf(xlim = c(-123.1, -122.4), ylim = c(45.3, 45.7)) + 
  scale_color_gradient(low = 'lightpink', high = 'purple') + 
  labs(title = "Security incidents, May-Jul 2022") + 
  scale_x_continuous(limits = c(-123.1, -122.4), expand = c(0, 0)) +
  scale_y_continuous(limits = c(45.3, 45.7), expand = c(0, 0))
```

Does this observation hold during April, September, and October (fewer than expected incidents?)
```{r}
df_apr_sep_oct_2022 = df3 %>% filter(month %in% c("Apr", "Sep", "Oct")) %>% group_by(location_id) %>% summarize(n = n()) %>% arrange(desc(n))   
  
  
df_apr_sep_oct_2022=  inner_join(df_apr_sep_oct_2022, df3 %>% select(location_id,stop_name, lon, lat, tractce, geometry.y) %>% distinct(), by = 'location_id')

ggplot() +
  geom_sf(data = tri_county_tracts) +
  geom_point(data = df_apr_sep_oct_2022, aes(x = lon, y = lat, size = n, color = n, alpha = n)) + 
  theme(legend.position = "none") + 
  coord_sf(xlim = c(-123.1, -122.4), ylim = c(45.3, 45.7)) + 
  scale_color_gradient(low = 'red', high = 'darkred') + 
  labs(title = "Security incidents, April, Sep, Oct 2022")


```


Choropleth to show hot spot areas in the city during the summer of 2022: 

```{r}
incident_per_tract_mjj_2022= df3 %>% filter(year == 2022 & month %in% c("May", "Jun", "Jul")) %>% group_by(tractce) %>%
  count() %>% left_join(df3 %>% select(tractce, namelsadco, geometry.y), by = "tractce") 

choropleth = function(subset){
ggplot(subset) +
  geom_sf(aes(geometry = geometry.y, fill = log(n))) + # logging n to condense the scale to show more variation 
  scale_fill_gradient(
      low = "white",
      high = "darkred",
      na.value = "grey50") +
  coord_sf(xlim = c(-123.1, -122.4), ylim = c(45.3, 45.7)) + 
  labs(title = "Trimet Security Incident Hotspots, May - July 2022") 
}

choropleth(incident_per_tract_mjj_2022)
```



Map of all security incidents in 2022: 

```{r}
ggplot() +
  geom_sf(data = tri_county_tracts) +
  geom_point(data = df3 %>% filter(year == 2022), aes(x = lon, y = lat), color = 'red', alpha = .05, size =.5) + 
  coord_sf(xlim = c(-123.1, -122.4), ylim = c(45.45, 45.6), expand = TRUE) + 
  theme(legend.position = "none")
 
```

Interactive Choropleth of 2022 data using leaflet 
```{r}
library(leaflet)
library(tidyverse)


leaflet_data= incident_per_tract_2022 %>% distinct() %>% st_as_sf()

pal = colorBin("YlOrRd", domain= leaflet_data$n, bins = c(0,50,100,150,200,250,300,350,400,450,500))

labels = sprintf(
  "<strong>%s</strong><br/>%g incidents",
  leaflet_data$tractce, leaflet_data$n
) %>% lapply(htmltools::HTML)

leaflet() %>%
  addProviderTiles("OpenStreetMap") %>% 
  addPolygons(data = leaflet_data, 
              fillColor = ~pal(n), 
              weight = 1,  
              opacity = 1,  
              fillOpacity = 0.5, 
              highlightOptions = highlightOptions(
                weight = 5, 
                color = "#666"), 
              label = labels, 
              labelOptions = labelOptions(style = list("font-weight" = "normal", padding = "3px 8px"),
                textsize = "15px",
                direction = "auto"))
```


```{r}

incident_per_tract_2022_2= df3 %>% filter(year == 2022) %>% group_by(tractce) %>%
  count() %>% left_join(df3 %>% select(tractce, namelsadco, geometry.y), by = "tractce") 


library(cartogram)

carto_data = incident_per_tract_2022_2 %>% distinct()  %>%  st_as_sf() 

carto_data = st_transform(carto_data, 3857)

cartog_cont = cartogram_cont(carto_data, weight = "n")

ggplot(cartog_cont) + geom_sf(aes(fill = n)) + labs(title = "Cartogram of Tracts with most Incidents in 2022") # Gateway, Rose Quarter, Downtown 
```


## Gateway Transit center is by far the place with most security incidents in 2022. 

What stops are the other 4 stops with the most security incidents in 2022? 
```{r}
top5_2022 = metro_2022 %>% arrange(desc(n)) %>% head(n =5) #10864 had 350 incidents, by far the most 

inner_join(top5_2022,df3 %>% select(location_id,location), by = 'location_id') %>% group_by(location_id,n) %>% distinct(location) %>% tail(n =5)
```
The other stops with the most security incidents are: 

Elmonica / SW 170th Ave MAX Station 
Hollywood/NE 42nd Ave TC 
Beaverton TC MAX Station 
Rose Quarter TC MAX Station

It does make sense that the MAX would have more security incidents, as they hold more people, often have less supervision on the cars, and they offer shelter / seclusion. It will be interesting to see what _types_ of incidents occur the most at these sites. 

Is this trend for Gateway TC true for other years? Does it match the overall trend in the system in the same time frame? 

```{r}
gateway_tc = df3 %>% filter(location_id == 10864)
gateway_tc %>% group_by(year) %>% count() %>% ggplot() + geom_col(aes(x = year, y = n)) # this is a familiar pattern...

df3 = df3 %>% mutate(is_gateway = ifelse(location_id == 10864, 1,0))

df3 %>% group_by(year, is_gateway) %>% count() %>% ggplot() + geom_col(aes(x = year, y = n, fill = factor(is_gateway, levels = c(0,1)))) + 
  guides(fill = guide_legend(title = "Is Gateway TC"))


```
The Gateway Transit Center is the leader in incidents, and the distribution follows the overall pattern of incidents across time. 

```{r}
gateway_year_counts = gateway_tc  %>% group_by(year) %>% count()       
df3_year_counts = df3 %>% filter(is_gateway == 0) %>% group_by(year) %>% count()

cor(gateway_year_counts$n,df3_year_counts$n) #
```

There is a strong correlation between security incidents across metro area and incidents at Gateway TC. 

In 2022, which tracts had more than 100 security incidents in that year ? 

```{r}
top15_tracts_2022 = incident_per_tract_2022 %>% filter(n >= 100) %>% distinct(tractce) %>% inner_join(df3 %>% filter(year == 2022), top15_tracts_2022, by = "tractce") 

#let's plot these events on the map 
ggplot() +
  geom_sf(data = tri_county_tracts) +
  geom_point(data = top15_tracts_2022 %>% group_by(location_id) %>% summarize(n = n(), lon, lat) %>% distinct(n, lon, lat), 
             aes(x = lon, y = lat, size = n), color = 'blue', alpha = .5) +
  coord_sf(xlim = c(-122.85, -122.4), ylim = c(45.425, 45.6)) + 
  theme(legend.position = "none") 

#this basically shows where each max line stops.
```
Top 15 tracts 
```{r}
unique(top15_tracts_2022$tractce)
```



It appears that it is the Red and Blue MAX lines play a big role in the incident numbers. Within these tracts, if we take out the MAX related security incidents, are they still high-incident areas? 
```{r}
df3 %>% filter(tractce %in% top15_tracts_2022$tractce & year == 2022 & type != 'MAX') %>% group_by(tractce) %>% count() %>% arrange(desc(n))

df3 %>% filter(tractce %in% top15_tracts_2022$tractce & year == 2022 ) %>% group_by(tractce) %>% count() %>% arrange(desc(n))

```
How many different routes intersect around Gateway TC ? 
```{r}
df3 %>% filter(tractce == '008100') %>% group_by(route_number) %>% summarize(n = n()) %>% drop_na() %>% nrow()
```




# We recommend examining the characteristics of the security incidents at Gateway Transit Center. It would appear that by proposing interventions at this location in the transit system, we could see the greatest effect on reducing security incidents.

Gateway TC is within tract 008100
The Gateway TC `location_id` is 10864

```{r}
df3 %>% filter(tractce == '008100' & year == 2022)
```

### Connections
MAX Blue Line
MAX Green Line
MAX Red Line
15-Belmont/NW 23rd
19-Woodstock/Glisan
22-Parkrose
23-San Rafael
24-Fremont
25-Glisan/Rockwood
87-Airport Way/181st
Columbia Area Transit
Columbia Gorge Express (Seasonal)

### Amenities
Park & Ride
Car sharing
Accessible to people with mobility devices
Ticket vending machine (credit/debit cards accepted)
Bike racks and lockers available
10-minute “Quick Drop” parking spaces for easy loading and unloading of passengers
Transit Tracker™ sign with MAX arrival countdown
Coffee Max: coffee, hot dogs and beverages during weekday commute hours


Which routes had the most security incidents in 2022 ? 
```{r}
unique(df3$route_number)

df3 %>% filter(type == 'MAX') %>% group_by(route_number) %>% count() %>% arrange(desc(n))


```

I cannot locate which routes (100, 200, 190, 925) are. I presume that these three digit route numbers are an internal code for the Max lines. I can filter the dataset and map the route stops to see which ones they are.  

```{r}
route_mapper = function(route){
  ggplot() +
  geom_sf(data = tri_county_tracts) +
  geom_point(data = df3 %>% filter(route_number == route), aes(x = lon, y = lat), color = 'red', alpha = .05, size =.5) + 
  coord_sf(xlim = c(-123.1, -122.4), ylim = c(45.45, 45.6), expand = TRUE) + 
  theme(legend.position = "none")
}
```

route 100 = BLUE LINE
```{r}
route_mapper(100)
```
route 200 = GREEN LINE 
```{r}
route_mapper(200)
```
route 190 = YELLOW LINE 
```{r}
route_mapper(190)
```

route 925 = Inconclusive
```{r}
route_mapper(925)
```
route 90 = RED 
```{r}
route_mapper(90)
```

route 290 = ORANGE LINE 
```{r}
route_mapper(290)
```
Adding Max color col to dataset 
```{r}
df3 = df3 %>% mutate(max_line = factor(case_when(
                              route_number == 100 ~ "blue", 
                              route_number == 200 ~ "green", 
                              route_number == 190 ~ "yellow", 
                              route_number == 90 ~ "red", 
                              route_number == 290 ~ "orange")))
```


## TYPE & SUBTYPE COLUMNS 
```{r}
subtype_breakdown = df3 %>% group_by(incident_subtype_code_list) %>% count() %>% arrange(desc(n)) # yikes, 37,436 are categorized as simply "NONE"

37436/nrow(df3) #78 % of these incidents do not have a subtype, this is marginally better than the 82% OTHER subtype in first dataset 


```


Resolving `NONE` issue 
```{r}
df3 = df3 %>% mutate(incident_subtype_code_list = gsub("NONE","",incident_subtype_code_list)) %>% #removing the word "NONE"
              mutate(incident_subtype_code_list = gsub("^\\s*$", NA, incident_subtype_code_list)) #changing empty rows to NA 

#let's get rid of those pesky commas 
df3 = df3 %>% mutate(incident_subtype_code_list = gsub(",","",incident_subtype_code_list)) %>% 
  mutate(incident_subtype_code_list = gsub("\\s", NA, incident_subtype_code_list)) #some more whitespace to get rid of 

unique(df3$incident_subtype_code_list)
```


What are the main subtypes we want to condense this subtype codes to ? 
```{r}
df3 %>% group_by(incident_subtype_code_list) %>% count()

```

Where are the fights happening ? 

Downtown, Rose Quarter, and along the east side of the MAX Blue Line 

```{r}
ggplot() +
  geom_sf(data = tri_county_tracts) +
  geom_point(data = df3 %>% filter(incident_subtype_code_list == 'FIGHT') , aes(x = lon, y = lat), color = 'red', alpha = .05, size =.5) + 
  coord_sf(xlim = c(-122.7, -122.5), ylim = c(45.45, 45.6), expand = TRUE) + 
  theme(legend.position = "none",
        )
```
How about Vandalism 
```{r}
ggplot() +
  geom_sf(data = tri_county_tracts) +
  geom_point(data = df3 %>% filter(incident_subtype_code_list == 'VANDAL'), aes(x = lon, y = lat), color = 'red', alpha = .05, size =.5) + 
  coord_sf(xlim = c(-123.1, -122.4), ylim = c(45.45, 45.6), expand = TRUE) + 
  theme(legend.position = "none")

df3_known_subtypes = df3 %>% filter(!is.na(incident_subtype_code_list))
```

Weapon-related incidents: 
```{r}
ggplot() +
  geom_sf(data = tri_county_tracts) +
  geom_point(data = df3 %>% filter(incident_subtype_code_list == 'TRESPASS') , aes(x = lon, y = lat), color = 'red', alpha = .05, size =.5) + 
  coord_sf(xlim = c(-123.1, -122.4), ylim = c(45.45, 45.6), expand = FALSE) + 
  theme(legend.position = "none") 
```

During what hour do most security incidents occur on the blue line ? 7 pm 

```{r}
df3 %>% filter(max_line == "blue") %>% 
  group_by(hour) %>% summarize(n = n()) %>% ggplot(aes(x = hour, y = n)) + geom_line(color = "blue") + labs(title = "The most incidents occur between 7 and 8 pm")
```

Does this change according to season? 
```{r}
df3 = df3 %>% mutate(season = case_when(
    month %in% c("Dec", "Jan", "Feb") ~ "Winter",
    month %in% c("Mar", "Apr", "May") ~ "Spring",
    month %in% c("Jun", "Jul", "Aug") ~ "Summer",
    month %in% c("Sep", "Oct", "Nov") ~ "Autumn",
    TRUE ~ as.character(month)
  ))

df3 %>% filter(max_line == "blue") %>% 
  group_by(hour, season) %>% summarize(n = n()) %>% ggplot(aes(x = hour, y = n)) + geom_line(color = "blue") + facet_wrap(~season) + labs(title = "Incidents peak between 5 and 8 pm on the MAX Blue Line")
```
Looks like evening, 5 - 8 pm is when the most activity occurs on MAX Blue line. 

How about for the whole system? 
```{r}
df3 %>% group_by(hour, season) %>% summarize(n = n()) %>% ggplot(aes(x = hour, y = n)) + geom_line(color = "black") + facet_wrap(~season)
```
Do MAX lines have different peak incident hours? 

```{r}
df3 %>% filter(!is.na(max_line)) %>% 
  group_by(hour, max_line) %>% summarize(n = n()) %>% ggplot(aes(x = hour, y = n)) + geom_line(aes(color = max_line)) + facet_wrap(~max_line) + scale_color_manual(values = c("blue", "green", "orange", "red", "yellow"))
```
 



Focusing on security incidents in 1 tract - lets do downtown 010602 

```{r}
tract_010602_2022 = df3 %>% filter(year == 2022 & tractce == '010602') %>% select(-geometry.x) %>% st_as_sf() 

leaflet() %>%
  addProviderTiles("OpenStreetMap") %>% 
  addPolygons(data = tract_010602_2022, 
              fillColor = "none") %>% 
  addMarkers(data = tract_010602_2022, lng = ~lon, lat  = ~lat)

?addMarkers()

tract_010602_2022$lon

```




