---
title: "trimet_analysis_v2.2"
author: Karol Orozco, Corey Cassell, Justus Eaglesmith, Charles Hanks, & CorDarryl Hall
output: html_document
date: "2023-07-05"
---

I. Time Analysis 
II. Location Analysis 
III. Incident Analysis 

FINDINGS: 

*  2022 had significantly higher security incidents than the previous years, with 36% more incidents than in 2021. 
*  Within that year, May through July has more activity, and April, Sep, and Oct had fewer incidents. 
*  Given the convergence of multiple routes, transit centers are hotspots for security incidents
*  Among the transit centers, Gateway Transit Center is a consistent hot spot, followed by Rose Quarter TC
*  During each day, we see the most security incidents during the hours of 4 to 8 pm.
*  Among all routes, the MAX Blue line is the most dangerous. 
*  Between the hours of 8 pm and 5 am, downtown (tract 010001) has the most security incidents 
*  1 in 3 incidents involve violence, intimidation, or abuse



Libraries: 
```{r}
library(tidyverse)
library(lubridate)
library(gridExtra)
library(forcats)
library(stringr)
library(tm)
library(SnowballC)
library(RColorBrewer)
library(tigris)
library(sf)
library(skimr)
library(leaflet)
library(leaflet.extras)
library(gifski)
library(gganimate)
library(lubridate)
library(ggmap)
```

Dataset: 
```{r}
df3 = read_rds('https://raw.githubusercontent.com/karolo89/capstone/main/data/df_spatial_clean.rds')
```

Need to standardize location names per location_id
```{r}
df3 = df3 %>% mutate(location = case_when(location_id == 10864 ~ "Gateway/NE 99th Ave TC MAX Station",
                                    location_id == 9654 ~ "Beaverton Transit Center",
                                    location_id == 9848 ~ "Hatfield Government Center MAX Station",
                                    location_id == 13248 ~ "Clackamas Town Center",
                                    location_id == 10856 ~ "Parkrose Sumner Transit Center",
                                    location_id == 10857 ~ "Gresham Central Transit Center",
                                    location_id == 8355 ~ "Ruby Junction/E 197th Ave MAX Station",
                                    TRUE ~ as.character(location)))

df3 = df3 %>% mutate(location = ifelse(location %in% c("Beaverton Transit Center","Beaverton TC MAX Station"), "Beaverton Transit Center", location))

```

# I.  TIME ANALYSIS 

First we examine how many security incidents occurred in each year and compare that the overall average per year. 

```{r}
df_year_counts <- df3 %>%
  count(year) %>%
  mutate(avg = mean(n),
        yoy = (n / lag(n) - 1) * 100, # adding year-over-year change 
         color = ifelse(n > avg, "Above Average", "Below Average"))

ggplot(df_year_counts, aes(x = year, y = n, fill = color)) + 
  geom_bar(stat = "identity", show.legend = FALSE) +
  geom_hline(aes(yintercept = avg), linetype = "dashed", color = "black") +
  scale_fill_manual(values = c("Above Average" = "#FF7F7F", "Below Average" = "#7F7FFF")) +
  theme_minimal() +
  labs(x = "Year", y = "Count", fill = "", title = "Number of Security Incidents Per Year") + 
  geom_label(data = df_year_counts %>% filter(year == 2022), aes(x = year, y = n, label = paste0(round(yoy, 2), "% +")), nudge_x = -1, fill = "yellow")
    
?geom_label()
```

The incidents are increasing annually so it is important to get a handle on this. Especially considering 2022 had tremendously more results than 2021. 

A better plot for this would be using a timeseries plot. 

```{r}
# Group data by year only and count the number of incidents
df_grouped <- df3 %>%
  group_by(year) %>%
  summarise(total_incident_count = n(), .groups = "drop")

# Create ggplot object
p <- ggplot(df_grouped, aes(x = year, y = total_incident_count)) +
  geom_line(size = 1, color = "blue") +
  geom_point(data = subset(df_grouped, year == 2022), colour = "red", size = 4, shape = 4) +
  labs(
    title = "Change in Total Number of Incidents Over Time",
    x = "Year",
    y = "Total Number of Incidents"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_y_log10() +
  geom_text(
    data = subset(df_grouped, year == 2022),
    aes(label = ifelse(year == 2022, as.character(total_incident_count), "")),
    vjust = -2.5,
    hjust = 0.5,
    size = 4,
    colour = "red"
  )

# Add animation
p_animate <- p + transition_reveal(year)

# Render the animation
animate(p_animate, end_pause = 50, height = 400, width = 600)

```

Clearly significant results in the year 2022. This will be the year we focus on to prime our analysis. Lets look at by location per year.

# By Location
```{r}
df_grouped <- df3 %>%
  group_by(year, location) %>%
  summarise(incident_count = n(), .groups = "drop")

df_grouped <- df_grouped %>%
  filter(location %in% top_locations)

#Find out the locations with the highest incidents in 2022
top_locations_2022 <- df_grouped %>%
  filter(year == 2022) %>%
  top_n(3, incident_count) %>%
  pull(location)

#Add a new variable to the dataframe representing the highlight
df_grouped <- df_grouped %>%
  mutate(highlight = ifelse(location %in% top_locations_2022, location, "Other"))

#Create a factor with the levels ordered to ensure "Other" is first (and therefore gray)
df_grouped$highlight <- factor(df_grouped$highlight, levels = c("Other", top_locations_2022))

#Create ggplot object
p <- ggplot(df_grouped, aes(x = year, y = incident_count, group = location, color = highlight)) +
  geom_line(size = 1) +
  geom_point(data = subset(df_grouped, year == 2022), size = 4, shape = 4) +
  scale_color_manual(values = c("gray", "red", "blue", "green")) +
  labs(
    title = "Change in Number of Incidents Over Time",
    x = "Year",
    y = "Number of Incidents",
    color = "Location"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_y_log10() +
  geom_text(
    data = subset(df_grouped, year == 2022),
    aes(label = ifelse(year == 2022, as.character(incident_count), "")),
    vjust = -.5,
    hjust = 0.5,
    size = 4,
    colour = "red"
  )

#Add animation
p_animate <- p + transition_reveal(year)

#Render the animation
animate(p_animate, end_pause = 50, height = 400, width = 600)
```





Appears that most of the incidents are occurring at Gateway, and Rose Quarter last year with some deviation between the top 3-5 following. Lets look to see if there is a trend with specific seasons/months.

```{r}
# Filter original dataframe to include only rows from the year 2022
df_2022 <- df3 %>%
  filter(year == 2022)

# Assign a 'season' column to df_2022
df_2022 <- df_2022 %>%
  mutate(season = case_when(
    month %in% c(12, 1, 2) ~ "Winter",
    month %in% 3:5 ~ "Spring",
    month %in% 6:8 ~ "Summer",
    month %in% 9:11 ~ "Autumn",
    TRUE ~ as.character(month)
  ))


# Assign a 'season' column to df_2022 based on 'month'
df_2022 <- df_2022 %>%
  mutate(season = case_when(
    month %in% c("Dec", "Jan", "Feb") ~ "Winter",
    month %in% c("Mar", "Apr", "May") ~ "Spring",
    month %in% c("Jun", "Jul", "Aug") ~ "Summer",
    month %in% c("Sep", "Oct", "Nov") ~ "Autumn",
    TRUE ~ as.character(month)
  ))
```

```{r}

# Calculate the number of incidents per month
df_sorted_2022 <- df_2022 %>%
   group_by(month, season) %>%
   summarise(incident_count = n(), .groups = "drop") %>%
   arrange(month)

# Calculate the average number of incidents per month
df_avg_2022 <- df_2022 %>%
  group_by(month) %>%
  summarise(avg_incident_count = n(), .groups = "drop") %>%
  arrange(month)

# Plotting the bar plot with the trendline
# Define a color palette
color_palette <- c("Winter" = "#1f78b4", "Spring" = "#33a02c", "Summer" = "#e31a1c", "Autumn" = "#ff7f00")

ggplot(df_sorted_2022, aes(x = month, y = incident_count, fill = season)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = color_palette) +
  labs(title = "Number of Incidents by Month in 2022", x = "Month", y = "Number of Incidents", fill = "Season") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "bottom") +
  geom_line(data = df_avg_2022, aes(x = month, y = avg_incident_count, group = 1), 
            color = "black", size = 1, inherit.aes = FALSE)
```

Appears to be a trend that in the warmer seasons incidents increase. Lets see if the statistics support this with a chi squared test of proportional independence.

```{r}

# Chi-square test for independence between month and incidents
chisq_test_month <- chisq.test(table(df_2022$month))

print(chisq_test_month)

# Chi-square test for independence between season and incidents
chisq_test_season <- chisq.test(table(df_2022$season))

print(chisq_test_season)

```
```{r}
# Extract standardized residuals
stdres <- chisq_test_month$stdres
names(stdres) <- month.abb

# Print residuals
print(stdres)

# Highlight residuals greater than 2 in absolute value
significant_months <- names(stdres)[abs(stdres) > 2]

print(paste("Months with significantly different number of incidents than expected: ", paste(significant_months, collapse = ", ")))

```

The following months show statistical significance, specifically in that April, September, and October have less than expected incidents while Jan, & May-July have more than expected incidents. Lets see if there are trends within time of day.


# II. LOCATION ANALYSIS 

Our analysis of incidents in 2022 during the months of May through July, there were higher than expected security incidents. Let us examine where these events occurred. 

Loading spatial data of Multnomah, Clackamas, and Washington counties:
```{r}
options(tigris_use_cache = TRUE)

this.year = 2020

mult_tracts = tracts(state = 'OR', county = "Multnomah", cb = T, year = this.year)
clack_tracts = tracts(state = 'OR', county = "Clackamas", cb = T, year = this.year)
wash_tracts = tracts(state = 'OR', county = "Washington", cb = T, year = this.year)

tri_county_tracts = bind_rows(mult_tracts,clack_tracts,wash_tracts)
```
Looking at where incidents occurred during peak period: May - July, 2022: 

```{r}
df_may_jul_2022 = df3 %>% 
  filter(date >= "2022-01-05" & date <= "2022-07-31") %>% 
  group_by(location, location_id) %>% 
  summarize(n = n()) %>% arrange(desc(n))

df_may_jul_2022 = inner_join(df_may_jul_2022, df3 %>% 
                  dplyr::select(location_id, lon, lat, tractce, geometry.y) %>% 
                  distinct(), by = 'location_id')

library(ggmap)
library(osmdata)

api_key = "AIzaSyBd4Q6cJzX1fFUhW2qgv6eWOztnZOcTuRk"
register_google(key = api_key)

portland_map = get_map(location = 'portland, or, usa', maptype = "roadmap", source = 'google', zoom = 10)

#bubble plot: 

ggmap(portland_map) +
  geom_point(data = df_may_jul_2022, aes(x = lon, y = lat, size = n, color = n, alpha = n)) + 
  theme(legend.position = "none") + 
  coord_sf(xlim = c(-123.1, -122.4), ylim = c(45.3, 45.7)) + 
  scale_color_gradient(low = 'red', high = 'darkred') + 
  labs(title = "Security incidents, May-Jul 2022")
```
In addition to Gateway, Hollywood, and the Rose Quarter, we can see that the Downtown Transit Mall has a high concentration of security incidents during this time period. 



Does this observation hold during April, September, and October, where there were fewer than expected incidents?
```{r}
df_apr_sep_oct_2022 = df3 %>% filter(year == 2022 & month %in% c("Apr","Sep","Oct")) %>% group_by(location, location_id) %>% summarize(n = n()) %>% arrange(desc(n))   
  
  
df_apr_sep_oct_2022=  inner_join(df_apr_sep_oct_2022, df3 %>% dplyr::select(location_id,stop_name, lon, lat, tractce, geometry.y) %>% distinct(), by = 'location_id')

ggmap(portland_map) +
  geom_point(data = df_apr_sep_oct_2022, aes(x = lon, y = lat, size = n, color = n, alpha = n)) + 
  theme(legend.position = "none") + 
  coord_sf(xlim = c(-123.1, -122.4), ylim = c(45.3, 45.7)) + 
  scale_color_gradient(low = 'red', high = 'darkred') + 
  labs(title = "Security incidents, April, Sep, Oct 2022")


```

Instead of view individual route stops, let's examine how security incident are distributed by area.  


Choropleth to show hot spot areas in the city during the summer of 2022: 

```{r}
incident_per_tract_mjj_2022= df3 %>% filter(year == 2022 & month %in% c("May", "Jun", "Jul")) %>% group_by(tractce) %>%
  count() %>% left_join(df3 %>% dplyr::select(tractce, namelsadco, geometry.y), by = "tractce") 

choropleth = function(subset){
ggplot(subset) +
  geom_sf(aes(geometry = geometry.y, fill = log(n))) + # logging n to condense the scale to show more variation 
  scale_fill_gradient(
      low = "white",
      high = "darkred",
      na.value = "grey50") +
  coord_sf(xlim = c(-123.1, -122.4), ylim = c(45.3, 45.7)) + 
  labs(title = "Trimet Security Incident Hotspots, May - July 2022") 
}

choropleth(incident_per_tract_mjj_2022)
```
We see a band of high-incident areas across the city, east to west. Given that the locations with the most incidents are at transit centers across this parallel, it appears that security incidents occur frequently around the Blue and Red MAX lines. 

Loading MAX data: 
```{r}
max_data = read_rds('https://raw.githubusercontent.com/karolo89/capstone/main/data/max_data.rds')

#subsetting each max line data
max_blue = max_data %>% filter(rte == 100 & dir == 0) %>% arrange(stop_seq)
max_red =  max_data %>% filter(rte == 90 & dir == 0) %>% arrange(stop_seq)
max_green = max_data %>% filter(rte == 200 & dir == 0) %>% arrange(stop_seq)
max_orange =  max_data %>% filter(rte == 290 & dir == 0) %>% arrange(stop_seq)
max_yellow = max_data %>% filter(rte == 190 & dir == 0) %>% arrange(stop_seq)

df3 = df3 %>% mutate(max_line = factor(case_when(
                              route_number == 100 ~ "blue", 
                              route_number == 200 ~ "green", 
                              route_number == 190 ~ "yellow", 
                              route_number == 90 ~ "red", 
                              route_number == 290 ~ "orange")))



```

Aggregating incident data for top 15 tracts: 

```{r}

incident_per_tract_2022 = df3 %>% 
  filter(year == 2022) %>% 
  group_by(tractce) %>%
  count() %>% 
  left_join(df3 %>% dplyr::select(tractce, namelsadco, geometry.y), by = "tractce") 


top15_tracts_2022 = incident_per_tract_2022 %>% 
  filter(n >= 100) %>% 
  distinct(tractce) %>% 
  inner_join(df3 %>% filter(year == 2022), top15_tracts_2022, by = "tractce") %>% 
  group_by(location_id) %>% 
  summarize(n = n(), lon, lat) %>% 
  distinct(n, lon, lat)


ggplot() +
  geom_sf(data = tri_county_tracts) +
  geom_point(data = top15_tracts_2022, 
             aes(x = lon, y = lat, size = n), color = 'blue', alpha = .5) +
  geom_path(data = max_red, aes(x = lon, y = lat), color = "red", size = 1, alpha = .5) + 
  geom_path(data = max_blue, aes(x = lon, y = lat), color = "blue", size = 1, alpha = .5) +
  geom_path(data = max_yellow, aes(x = lon, y = lat), color = "yellow", size = 1, alpha = .5) +
  geom_path(data = max_orange, aes(x = lon, y = lat), color = "orange", size = 1, alpha = .5) +
  geom_path(data = max_green, aes(x = lon, y = lat), color = "green", size = 1, alpha = .5) +
  labs(title =  "TriMet MAX System") +
  coord_sf(xlim=c(-123.1, -122.4), ylim = c(45.40,45.6)) +
  theme(legend.position = "none") 
```
From this visualization it is clear that the security incident hot spots are along the max routes. 


Which MAX routes had the most security incidents ? 
```{r}
df3 %>% filter(type == 'MAX') %>% group_by(route_number, max_line) %>% count() %>% arrange(desc(n))

```
During what hour do most security incidents occur on the blue line ?  

```{r}
df3 %>% filter(max_line == "blue") %>% 
  group_by(hour) %>% summarize(n = n()) %>% ggplot(aes(x = hour, y = n)) + geom_line(color = "blue") + labs(title = "The most incidents occur between 7 and 8 pm")
```
During this peak time, where are these incidents occuring?  

```{r}

blue_7pm = df3 %>% filter(max_line == "blue" & hour == 19) %>% count(location_id) %>% inner_join(df3 %>% dplyr::select(location_id,location, lon, lat), by = "location_id") %>% distinct(location_id, n, lon, lat) %>% arrange(desc(n)) %>% head(n = 5) %>% distinct(location_id, n, lon, lat)


ggplot() + 
  geom_sf(data = tri_county_tracts) +
  geom_point(data = blue_7pm, aes(x = lon, y = lat, size = n), color = 'red', alpha = .5, ) + 
  coord_sf(xlim = c(-123.1, -122.4), ylim = c(45.3, 45.6)) + 
  theme(legend.position = "none")



```
```{r}
blue_7p_tract = df3 %>% filter(max_line == "blue" & hour == 19) %>% count(tractce) %>% inner_join(df3 %>% dplyr::select(tractce, geometry.y), by = "tractce") %>% distinct(tractce, n, geometry.y)

choropleth(blue_7p_tract)

```


Does this change according to season? 
```{r}
df3 = df3 %>% mutate(season = case_when(
    month %in% c("Dec", "Jan", "Feb") ~ "Winter",
    month %in% c("Mar", "Apr", "May") ~ "Spring",
    month %in% c("Jun", "Jul", "Aug") ~ "Summer",
    month %in% c("Sep", "Oct", "Nov") ~ "Autumn",
    TRUE ~ as.character(month)
  ))

df3 %>% filter(max_line == "blue") %>% 
  group_by(hour, season) %>% summarize(n = n()) %>% ggplot(aes(x = hour, y = n)) + geom_line(color = "blue") + facet_wrap(~season) + labs(title = "Incidents peak between 5 and 8 pm on the MAX Blue Line")
```

Looks like evening, 5 - 8 pm is when the most activity occurs on MAX Blue line. 

How about for the whole system? 
```{r}
df3 %>% group_by(hour, season) %>% summarize(n = n()) %>% ggplot(aes(x = hour, y = n)) + geom_line(color = "black") + facet_wrap(~season)
```
Do MAX lines have different peak incident hours? 

```{r}
df3 %>% filter(!is.na(max_line)) %>% 
  group_by(hour, max_line) %>% summarize(n = n()) %>% ggplot(aes(x = hour, y = n)) + geom_line(aes(color = max_line)) + facet_wrap(~max_line) + scale_color_manual(values = c("blue", "green", "orange", "red", "yellow"))
```

Interactive Choropleth of 2022 data using leaflet 
```{r}
leaflet_data= incident_per_tract_2022 %>% distinct() %>% st_as_sf()

pal = colorBin("YlOrRd", domain= leaflet_data$n, bins = c(0,50,100,150,200,250,300,350,400,450,500))

labels = sprintf(
  "<strong>%s</strong><br/>%g incidents",
  leaflet_data$tractce, leaflet_data$n
) %>% lapply(htmltools::HTML)

leaflet() %>%
  addProviderTiles("OpenStreetMap") %>% 
  addPolygons(data = leaflet_data, 
              fillColor = ~pal(n), 
              weight = 1,  
              opacity = 1,  
              fillOpacity = 0.5, 
              highlightOptions = highlightOptions(
                weight = 5, 
                color = "#666"), 
              label = labels, 
              labelOptions = labelOptions(style = list("font-weight" = "normal", padding = "3px 8px"),
                textsize = "15px",
                direction = "auto")) %>% 
  addPolylines(data = max_blue, lng = ~lon, lat = ~lat, color = 'blue') %>% 
  addPolylines(data = max_red, lng = ~lon, lat = ~lat, color = 'red') %>% 
  addPolylines(data = max_yellow, lng = ~lon, lat = ~lat, color = 'yellow') %>% 
  addPolylines(data = max_orange, lng = ~lon, lat = ~lat, color = 'orange')  %>% 
  addPolylines(data = max_green, lng = ~lon, lat = ~lat, color = 'green') 


?addPolylines()
```


## Gateway Transit center is by far the place with most security incidents in 2022. 

What stops are the other 4 stops with the most security incidents in 2022? 
```{r}
top5_2022 = metro_2022 %>% arrange(desc(n)) %>% head(n =5) #10864 had 350 incidents, by far the most 

inner_join(top5_2022,df3 %>% dplyr::select(location_id,location), by = 'location_id') %>% group_by(location_id,n) %>% distinct(location) %>% tail(n =5)
```
The other stops with the most security incidents are: 

Elmonica / SW 170th Ave MAX Station 
Hollywood/NE 42nd Ave TC 
Beaverton TC MAX Station 
Rose Quarter TC MAX Station

It does make sense that the MAX would have more security incidents, as they hold more people, often have less supervision on the cars, and they offer shelter / seclusion. 

Is this trend for Gateway TC true for other years? Does it match the overall trend in the system in the same time frame? 

```{r}
gateway_tc = df3 %>% filter(location_id == 10864)
gateway_tc %>% group_by(year) %>% count() %>% ggplot() + geom_col(aes(x = year, y = n)) # this is a familiar pattern...

df3 = df3 %>% mutate(is_gateway = ifelse(location_id == 10864, 1,0))

df3 %>% group_by(year, is_gateway) %>% count() %>% ggplot() + geom_col(aes(x = year, y = n, fill = factor(is_gateway, levels = c(0,1)))) + 
  guides(fill = guide_legend(title = "Is Gateway TC"))


```
The Gateway Transit Center is the leader in incidents, and the distribution follows the overall pattern of incidents across time. 

```{r}
gateway_year_counts = gateway_tc  %>% group_by(year) %>% count()       
df3_year_counts = df3 %>% filter(is_gateway == 0) %>% group_by(year) %>% count()

cor(gateway_year_counts$n,df3_year_counts$n) #
```

There is a strong correlation between security incidents across metro area and incidents at Gateway TC. 


It appears that it is the Red and Blue MAX lines play a big role in the incident numbers. Within these 15 tracts, if we take out the MAX related security incidents, are they still high-incident areas? 
```{r}
df3 %>% filter(tractce %in% top15_tracts_2022$tractce & year == 2022 & type != 'MAX') %>% group_by(tractce) %>% count() %>% arrange(desc(n))

df3 %>% filter(tractce %in% top15_tracts_2022$tractce & year == 2022 ) %>% group_by(tractce) %>% count() %>% arrange(desc(n))

```
Even not counting max related security incidents - tract 008100 (Gateway) is a big hot spot. The other tracts' number decrease significantly without the max line incidents. 


How many different routes intersect around Gateway TC ? 
```{r}
df3 %>% filter(tractce == '008100') %>% group_by(route_number) %>% summarize(n = n()) %>% drop_na() %>% nrow()
```


## Next let us look at the characteristics of the security incidents at Gateway Transit Center. 

Gateway TC is within tract 008100
The Gateway TC `location_id` is 10864

```{r}
df3 %>% filter(tractce == '008100' & year == 2022)
```

### Connections
MAX Blue Line
MAX Green Line
MAX Red Line
15-Belmont/NW 23rd
19-Woodstock/Glisan
22-Parkrose
23-San Rafael
24-Fremont
25-Glisan/Rockwood
87-Airport Way/181st
Columbia Area Transit
Columbia Gorge Express (Seasonal)

### Amenities
Park & Ride
Car sharing
Accessible to people with mobility devices
Ticket vending machine (credit/debit cards accepted)
Bike racks and lockers available
10-minute “Quick Drop” parking spaces for easy loading and unloading of passengers
Transit Tracker™ sign with MAX arrival countdown
Coffee Max: coffee, hot dogs and beverages during weekday commute hours





## TYPE & SUBTYPE COLUMNS 
```{r}
subtype_breakdown = df3 %>% group_by(incident_subtype_code_list) %>% count() %>% arrange(desc(n)) # yikes, 37,436 are categorized as simply "NONE"

37436/nrow(df3) #78 % of these incidents do not have a subtype, this is marginally better than the 82% OTHER subtype in first dataset 


```


Resolving `NONE` issue 
```{r}
df3 = df3 %>% mutate(incident_subtype_code_list = gsub("NONE","",incident_subtype_code_list)) %>% #removing the word "NONE"
              mutate(incident_subtype_code_list = gsub("^\\s*$", NA, incident_subtype_code_list)) #changing empty rows to NA 

#let's get rid of those pesky commas 
df3 = df3 %>% mutate(incident_subtype_code_list = gsub(",","",incident_subtype_code_list)) %>% 
  mutate(incident_subtype_code_list = gsub("\\s", NA, incident_subtype_code_list)) #some more whitespace to get rid of 

unique(df3$incident_subtype_code_list)

df3 %>% count(incident_subtype_code_list)

39934/48201
```


What are the main subtypes we want to condense this subtype codes to ? 
```{r}
df3 %>% group_by(incident_subtype_code_list) %>% count() %>% arrange(desc(n))

df3 %>% filter(year == 2022) %>% filter(!is.na(incident_subtype_code_list)) %>% group_by(incident_subtype_code_list) %>% count() %>% arrange(desc(n))
```

Where are the fights happening ? 

Downtown, Rose Quarter, and along the east side of the MAX Blue Line 

```{r}
ggplot() +
  geom_sf(data = tri_county_tracts) +
  geom_point(data = df3 %>% filter(incident_subtype_code_list == 'FIGHT') , aes(x = lon, y = lat), color = 'red', alpha = .05, size =.5) + 
  coord_sf(xlim = c(-122.7, -122.5), ylim = c(45.45, 45.6), expand = TRUE) + 
  theme(legend.position = "none",
        )
```
How about Vandalism 
```{r}
ggplot() +
  geom_sf(data = tri_county_tracts) +
  geom_point(data = df3 %>% filter(incident_subtype_code_list == 'VANDAL'), aes(x = lon, y = lat), color = 'red', alpha = .05, size =.5) + 
  coord_sf(xlim = c(-123.1, -122.4), ylim = c(45.45, 45.6), expand = TRUE) + 
  theme(legend.position = "none")

df3_known_subtypes = df3 %>% filter(!is.na(incident_subtype_code_list))
```

Trespass-related incidents: 
```{r}
ggplot() +
  geom_sf(data = tri_county_tracts) +
  geom_point(data = df3 %>% filter(incident_subtype_code_list == 'TRESPASS') , aes(x = lon, y = lat), color = 'red', alpha = .05, size =.5) + 
  coord_sf(xlim = c(-123.1, -122.4), ylim = c(45.45, 45.6), expand = FALSE) + 
  theme(legend.position = "none") 
```
Does it seem like trespassing mainly occurs around max lines ? 
```{r}
 df3 %>% filter(incident_subtype_code_list == 'TRESPASS') %>% group_by(route_number) %>% count()
```
Yes, probably because there are more restricted areas around the max rail. 


 
Focusing on security incidents in 1 tract - lets do downtown 010602 

```{r}
tract_010602_2022 = df3 %>% filter(year == 2022 & tractce == '010602') %>% dplyr::select(-geometry.x) %>% st_as_sf() 

leaflet() %>%
  addProviderTiles("OpenStreetMap") %>% 
  addPolygons(data = tract_010602_2022, 
              fillColor = "none") %>% 
  addMarkers(data = tract_010602_2022, lng = ~lon, lat  = ~lat)


```


Among the labeled subtypes, what sort of incidents occur at Gateway transit center, location ID 10864 ?

```{r}
gateway_incidents = df3 %>% filter(!is.na(incident_subtype_code_list)) %>% filter(year == 2022) %>%  filter(location_id == 10864) %>% group_by(incident_subtype_code_list) %>% count()

sum(gateway_incidents$n) # 459 total incidents 

gateway_incidents = gateway_incidents %>% arrange(desc(n)) %>%mutate(freq = n/43)
```
Trespassing, vehicle tow, vandalism, fights, and customer assault are the most common incidents at this stop. 


How does this compare with the rest of the system? 

```{r}
df3 %>% filter(!is.na(incident_subtype_code_list)) %>% group_by(incident_subtype_code_list) %>% count() %>% arrange(desc(n)) 

df3 %>% filter(year == 2022 & !is.na(incident_subtype_code_list))
```

Overall, the most frequent occurences are vandalism, trespressing, assault of customer, fight, and assault of employee.


# Analysis of the NIGHT: 

ASSUMPTION: The night is between 8 pm and 5 am: 
```{r}
df3 = df3 %>% mutate(night_time = factor(ifelse(hour >= 20 | hour <= 4, 1, 0))) #creating night time flag 

df3 %>% group_by(night_time, season) %>% count()  %>% 
            ggplot(aes(x = season, y = n, fill = night_time)) + geom_col() + facet_grid(~ night_time) + 
            scale_discrete_manual(aesthetics = c("fill"), values = c("#FF7F7F", "#7F7FFF")) + labs(title = "# of Incidents, Day vs Night Per Season")
```

```{r}
df3 %>% filter(year == 2023) %>% group_by(date, night_time) %>% count() %>%  ggplot(aes(x = date, y = n, fill = night_time))  + geom_point() + 
  geom_smooth(method = 'lm', se = FALSE) + 
  facet_grid(~night_time) + 
  labs(title = "The number of night time incidents are decreasing in 2023")
```

What if we compare the same time interval, one year ago? 
```{r}
df3 %>% filter(year %in% c(2022,2023))%>%
  filter((date >= "2022-01-01" & date <= "2022-05-16") | date >= "2023-01-01")%>% 
  group_by(date,year, night_time) %>% count() %>%  ggplot(aes(x = date, y = n, color = night_time))  + geom_point() + 
  geom_smooth(method = 'lm', se = FALSE, color = "black") + 
  facet_grid(night_time~year) + 
  theme_minimal() + labs(title = "Jan - May, 2022 vs 2023")
```
Nighttime incidents were trending upward in Jan - May 2022, and now they are trending downward Jan - May 2023. 

Let's look at the whole year of 2022 see how many incidents were occuring at night: 
```{r}
df3 %>% filter(year == 2022 & night_time == 1) %>% group_by(date) %>% count() %>% 
  ggplot(aes(x = date, y = n))  + geom_point() + 
  geom_smooth(method = "loess")
```

The most incidents occur between 4 and 8 pm
```{r}
df3 %>%
  group_by(hour) %>% summarize(n = n()) %>% ggplot(aes(x = hour, y = n)) + geom_line(color = "blue") + labs(title = "The most incidents occur between 7 and 8 pm")

df3 %>% group_by(hour) %>% summarize(n = n()) %>% arrange(desc(n))
```


What part of the system is most dangerous between the hours of 8 pm and 5 am ? 

Where are the most incidents occurring at night? Downtown, Rose Quarter, Gateway, Downtown 2

```{r}
df3 %>% filter(night_time == 1) %>%  
  group_by(tractce) %>% 
    count() %>% 
      arrange(desc(n))

```

# INCIDENT SUBTYPE CATEGORIZATION 

Objective: Classifying incidents based on words present in comments column: 


```{r}
df22 = df3 %>% filter(year == 2022)
```

Sleepers 

```{r}
df22 = df22 %>% mutate(sleeper = ifelse(str_detect(comments, "[Ss]leeper"), 1, 0))
df22 %>% filter(sleeper == 1) %>% group_by(location) %>% count() %>% arrange(desc(n))
```
The result of the sleeper class is too lean to do meaningfu analysis. Same with alcohol/drugs. 

DICTIONARIES 
```{r}
drug_words = c('[Dd]rug','[Ss]hooting up','[Ss]moke','[Vv]ape', '[Pp]araphenelia','[Bg]aggie','[Nn]eedle','[Hh]igh', '[Hh]erion','[Mm]eth', '[Pp]ipe', '[Cc]ocaine', '[Hh]ypodermic','[Ff]ume', '[Dd]ose','[Ff]entanyl','[Nn]arcotic','[Ii]nhalent', '[Cc]annabis','[Mm]arijuana','[Ff]oil')


df22 %>% filter(str_detect(comments, pattern = paste(drug_words, collapse = "|"))) #568 incidents containing one or more of the words above 

df22 = df22 %>% mutate(drugs = ifelse(str_detect(comments, pattern = paste(drug_words, collapse = "|")),1,0))



```

```{r}
alcohol_words = c('[Bb]eer','[Aa]lcohol','[Dd]runk', "[Ii]ntoxicat", "[Dd]k")

df22 %>% filter(str_detect(comments, pattern = paste(alcohol_words, collapse = "|")))  

df22 = df22 %>% mutate(alcohol = ifelse(str_detect(comments, pattern = paste(alcohol_words, collapse = "|")),1,0))
```
do not board (86'd people)
```{r}
df22 %>% filter(str_detect(comments, pattern = regex("do not board", ignore_case = TRUE)))

df22 = df22 %>% mutate(do_not_board = ifelse(str_detect(comments, pattern = regex("do not board", ignore_case = TRUE)),1,0))

```
```{r}
violent_words <- c('[Ss]wing', '[Hh]it', '[Ss]truck', '[Aa]ssault', '[Tt]hrow', '[Gg]rab',
           '[Tt]hreat', '[Pp]hysic', '[Mm]enac','[Ff]ace','[Ff]led', '[Ii]njur', '[Ss]teel pole', '[Ss]pit','[Ss]hov', '[Ss]lam','[Hh]urt', '[Bb]elig','[Pp]unch', '[Gg]oing off', '[Yy]ell','[Aa]ggressive','[Gg]un', '[Kk]nife','[Mm]achetee','[Aa]xe','[Br]andish','[Hh]achet','[Aa]xe','[Aa]ttack', '[Aa]ggres', '[Kk]ill', '[Pp]ound', '[Ss]uicid', '[Ff]ire', '[Ff]irearm', '[Hh]arras', '[Bb]lood')

df22 %>% filter(str_detect(comments, pattern = paste(violent_words, collapse = "|")))

df22 = df22 %>% mutate(violence = ifelse(str_detect(comments, pattern = paste(violent_words, collapse = "|")),1,0))

df22 %>% filter(violence == 1) %>% nrow()

2223/nrow(df22)

```

34% of incidents in 2022 could be considered danger/harassment/physical-verbal intimidation acts.


Ranking violent incidents by transit center/max stop in 2022:
```{r}
#top 100 locations where these sort of incidents occur in 2022
viol_locs_22 = df22 %>% filter(violence == 1 & str_detect(location, paste(c("MAX", "TC", "Center"), collapse = "|"))) %>% group_by(location) %>% count() %>% arrange(desc(n)) %>% rename(viol_inc_count = n)

#getting total incidents per these locations in 2022
total_incidents_locs22 = df22 %>% filter(location %in% viol_locs_22$location) %>% group_by(location) %>% count() %>% rename(total_inc_count = n)

#joining two aggregate tables to create percentage of violent acts by total acts 
#filter danger_zones by more than 20 incidents in 2022 
danger_zones = inner_join(total_incidents_locs22, viol_locs_22, by = "location") %>% mutate(perc_danger = viol_inc_count/total_inc_count) %>% filter(total_inc_count >= 20) %>% arrange(desc(perc_danger)) 


df22approx = df22 %>% mutate(lon = round(lon,2), 
               lat = round(lat,2))

#joining spatial data with subset: 
df22approx = left_join(danger_zones, df22approx %>% dplyr::select(location, lon, lat), by = "location") %>% distinct()

```

Among incidents at these locations, here are the highest ratio of danger incidents per total incidents: 

```{r}
danger_zones %>% head(n = 10) %>% ggplot(aes(x = reorder(location, perc_danger), y = perc_danger)) + geom_col() + coord_flip()
```

Plotting the danger zones: 
```{r}
ggplot() +
  geom_sf(data = tri_county_tracts) +
  geom_point(data = df22approx, aes(x = lon, y = lat, color = perc_danger, size = perc_danger)) +
  geom_path(data = max_red, aes(x = lon, y = lat), color = "red", size = 1, alpha = .3) + 
  geom_path(data = max_blue, aes(x = lon, y = lat), color = "blue", size = 1, alpha = .3) +
  geom_path(data = max_yellow, aes(x = lon, y = lat), color = "yellow", size = 1, alpha = .3) +
  geom_path(data = max_orange, aes(x = lon, y = lat), color = "orange", size = 1, alpha = .3) +
  geom_path(data = max_green, aes(x = lon, y = lat), color = "green", size = 1, alpha = .3) +
  scale_color_gradient(low = "white", high = "red") + 
  coord_sf(xlim=c(-123.1, -122.4), ylim = c(45.40,45.65)) + 
  theme(legend.position = "none")
```
### Term frequency: 

subtype VANDAL
```{r}
library(tidytext)
df3_vandalism = df3 %>% filter(incident_subtype_code_list == "VANDAL")

corpus <- Corpus(VectorSource(df3_vandalism$comments))
corpus <- tm_map(corpus, content_transformer(tolower))
corpus <- tm_map(corpus, removePunctuation)
corpus <- tm_map(corpus, removeNumbers)
corpus <- tm_map(corpus, removeWords, stopwords("english"))
corpus <- tm_map(corpus, stemDocument)

dtm <- DocumentTermMatrix(corpus)

tidy_dtm = tidy(dtm)

vandal_top_words <- tidy_dtm %>%
  group_by(term) %>%
  summarise(total = sum(count)) %>%
  arrange(desc(total)) %>% filter(!(term %in% c('sent', 'incid','email', 'group', 'datetim', 'busmanagerstrimetorg','securityvand','call','stop','report','oper','supervisor','request','will','person','polic','respond','passeng','check','busstationmanagerstrimetorg'))) %>% filter(total >= 100)
```
```{r}
vandal_top_words$term

#select vandalism words from top words list above: 
vandalism_words =  c("shelter", "glass", "broken", "window", "graffiti", "broke", "threw", "rock", "shatter", "board", "tape", "repair", "clean", "vandal", "fire", "break", "crack", "windshield", "repairbus", "spiderweb")
```


```{r}
df22 %>% filter(str_detect(comments, pattern = paste(vandalism_words, collapse = "|"))) %>% group_by(incident_subtype_code_list) %>% count() %>% arrange(desc(n))
```
This seems somewhat functional... 

subtype TRESPASS 
```{r}
library(tidytext)
df3_trespass = df3 %>% filter(incident_subtype_code_list == "TRESPASS")

corpus <- Corpus(VectorSource(df3_trespass$comments))
corpus <- tm_map(corpus, content_transformer(tolower))
corpus <- tm_map(corpus, removePunctuation)
corpus <- tm_map(corpus, removeNumbers)
corpus <- tm_map(corpus, removeWords, stopwords("english"))
corpus <- tm_map(corpus, stemDocument)

dtm <- DocumentTermMatrix(corpus)

tidy_dtm = tidy(dtm)

trespass_words <- tidy_dtm %>%
  group_by(term) %>%
  summarise(total = sum(count)) %>%
  arrange(desc(total)) %>% filter(!(term %in% c('sent', 'incid','email', 'group', 'datetim', 'busmanagerstrimetorg','securityvand','call','stop','report','oper','supervisor','request','will','person','polic','respond','passeng','check','busstationmanagerstrimetorg'))) %>% filter(total >= 200) %>% head(n = 10)
```


```{r}
df22 %>% filter(str_detect(comments, pattern = paste(trespass_words$term, collapse = "|"))) %>% group_by(incident_subtype_code_list) %>% count() %>% arrange(desc(n))

```
Not as good results. 

What if we are just to filter comments for word "trespass" 
```{r}
df22 %>% filter(str_detect(comments,"trespass")) %>% filter(is.na(incident_subtype_code_list))


```
Yes, this is far more effective. 

subtype = Fight vs. subtype ASLTCUST ? 
```{r}
df22 %>% filter(str_detect(comments, '[Ff]ight')) %>% group_by(incident_subtype_code_list) %>% count()
```
```{r}
df3_asltcust = df3 %>% filter(incident_subtype_code_list == "ASLTCUST")

corpus <- Corpus(VectorSource(df3_asltcust$comments))
corpus <- tm_map(corpus, content_transformer(tolower))
corpus <- tm_map(corpus, removePunctuation)
corpus <- tm_map(corpus, removeNumbers)
corpus <- tm_map(corpus, removeWords, stopwords("english"))
corpus <- tm_map(corpus, stemDocument)

dtm <- DocumentTermMatrix(corpus)

tidy_dtm = tidy(dtm)

asltcust_words <- tidy_dtm %>%
  group_by(term) %>%
  summarise(total = sum(count)) %>%
  arrange(desc(total)) %>% filter(!(term %in% c('sent', 'incid','email', 'group', 'datetim', 'busmanagerstrimetorg','securityvand','call','stop','report','oper','supervisor','request','will','person','polic','respond','passeng','check','busstationmanagerstrimetorg'))) %>% filter(total >= 200)
```

```{r}
df3_asltempl = df3 %>% filter(incident_subtype_code_list == "ASLTEMPL")

corpus <- Corpus(VectorSource(df3_asltempl$comments))
corpus <- tm_map(corpus, content_transformer(tolower))
corpus <- tm_map(corpus, removePunctuation)
corpus <- tm_map(corpus, removeNumbers)
corpus <- tm_map(corpus, removeWords, stopwords("english"))
corpus <- tm_map(corpus, stemDocument)

dtm <- DocumentTermMatrix(corpus)

tidy_dtm = tidy(dtm)

asltempl_words <- tidy_dtm %>%
  group_by(term) %>%
  summarise(total = sum(count)) %>%
  arrange(desc(total)) %>% filter(!(term %in% c('sent', 'incid','email', 'group', 'datetim', 'busmanagerstrimetorg','securityvand','call','stop','report','oper','supervisor','request','will','person','polic','respond','passeng','check','busstationmanagerstrimetorg'))) %>% filter(total >= 200)
```



